---
title: "LIVER CIRRHOSIS SURVIVAL ANALYSIS"
author: "Enock Bereka"
date: "2025-02-04"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries
```{r warning=FALSE, message=FALSE}
library(survminer)
library(survival)
library(flexsurv)
library(tidyverse)
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
data <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/survival/cirrhosis.csv")
glimpse(data)
anyNA(data)
data <- drop_na(data)
duplicated(data) %>% table()
```

## Convert character variables to factors
```{r warning=FALSE, message=FALSE}
data <- data %>% mutate_if(is.character, as.factor)
```

## What is the probability of survival over time for cirrhosis patients?
## How does survival differ between treatment groups (D-penicillamine vs. placebo)?
## Kaplan-Meier Model
```{r warning=FALSE, message=FALSE}
km_fit <- survfit(Surv(N_Days, Status == "D") ~ Drug, data = data)
km_fit
```

## We had 136 participants under D-penicillamine when the study began. 57 of the participants died from liver cirrhosis during the study.
## We had 140 participants under Placebo when the study began. 54 of them died from liver cirrhosis during the study.
## Half the participants under D-penicillamine survived beyond 3086 days half did not.
## Half the participants under Placebo survived beyond 3428 days half did not.

## Log-rank test
```{r warning=FALSE, message=FALSE}
log_rank <- survdiff(Surv(N_Days, Status == "D") ~Drug , data = data)
log_rank
```

## There was no significant difference in survival between those participants under Placebo and those under D-penicillamine 

## Plot Kaplan-Meier survival curves
```{r warning=FALSE, message=FALSE}
ggsurvplot(km_fit, data = data, 
           conf.int = FALSE, 
           pval = TRUE, 
           mark.time = T,
           risk.table = TRUE, 
           legend.title = "Treatment Group", 
           xlab = "Days", ylab = "Survival Probability",
           title = "Kaplan-Meier Survival Curves for Cirrhosis Patients",
           lower.tail = FALSE)
```

## The kaplan meier survival curve shows a clear stratification based on Drug, with those under D-penicillamine having slightly shorter survival time compared to those under Placebo

## Which factors significantly affect survival
## Fit Cox Proportional Hazards Model
```{r warning=FALSE, message=FALSE}
cox_fit <- coxph(Surv(N_Days, Status == "D") ~ Drug + Age + Sex + Bilirubin + Albumin + Ascites + Edema + Stage, data = data)
```

## Summarize results
```{r warning=FALSE, message=FALSE}
summary(cox_fit)
```

## There were 276 participants in the study and 111 particants died from Liver cirrhosis.
## Likelihood Ratio Test, Wald Test and Score Test
## These tests assesses whether the included covariates are statistically signicant.
## All tests yielded a p-value of 2e-16, explaining the covariates significantly explain the hazard rates and survival time.
## The model yielded a concordance of 0.834 suggesting it high predictive power.
## Participants under Placebo have 0.1076 lower risk of death compared to those under D-penicillamine. The p-value is statistically insignificant.
## For each unit increase in Age, the hazard rate decreases by 0.01%. The p-value is statistically significant.
## Males have 90.73 higher risk of death compared to female.
## For each unit increase in Bilirubin , the hazard rate increases by 12.94%
## For each unit increase in Albumin , the hazard rate decreases by 59.23%
## Patients with Ascites have 11.89 higher risk of death compared to those without.
## Patients with EdemaS have 15.90 higher risk of death compared to those without.
## Patients with EdemaY have 57.45 higher risk of death compared to those without.
## For each unit increase in stage, the hazard rate increases by 61.26%.

## Check proportional hazard assumption.
```{r warning=FALSE, message=FALSE}
ph_tet <- cox.zph(cox_fit)
print(ph_tet)
```

## The PH assumption holds globally, suggesting the model is valid. The predictors explain hazard variation modestly well.

## Visualize model assumptions
```{r warning=FALSE, message=FALSE}
ggcoxzph(cox.zph(cox_fit))
```

## Visual inspection of Schoenfeld residual plots  confirms that the PH assumption is not violated.

## Visualize the survival curves.
```{r warning=FALSE, message=FALSE}
ggsurvplot(survfit(cox_fit), data = data, risk.table = TRUE)
```

